Supporting Information:\\Black-swan events in animal populations

### 2016-06-25 text/ms.tex

\section{Black-swan events in animal populations}

@@

We chose the $t$ distribution for a number of reasons. 
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes 
that could give rise to extreme events,
such as spatial heterogeneity, nonstationary or heavy-tailed environmental processes, 
or varying responses to temperature among individuals. 
While evaluating these various hypotheses 
would be important for diagnosing the underlying causes 
of extreme events and to develop predictive models, 
it would be impossible to tease out each of these 
underlying causes from population trend data alone in a comparative manner. 
Instead we seek a simple flexible phenomenological model 
that can be applied across the widest possible 
range of population time series to provide a 
comparative overview of the prevalence 
of black-swan events.

One alternative, the generalized extreme value (GEV) distribution,
represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold.
Therefore, the GEV requires longer time series then typically available
for wild populations.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

@@

\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals: \begin{align} x_t &= \lambda + b x_{t-1} + \epsilon_t\\ \epsilon_t
&\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma). \end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \citep{connors2014} and the chosen prior in previous
analyses \citep{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was systematically affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

@@

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{ln} abundance.

@@

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from \citet{brook2006a}.
The beta distribution is useful when response data are on a continuous scale
between zero and one \citep{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
10)\) on the global intercept \(\alpha\), \(\mathrm{Half\mhyphen Cauchy}(0,
2.5)\) on all standard deviation parameters, \(\mathrm{Half\mhyphen Cauchy}(0,
10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
probability density around expected parameter values while allowing for a
higher probability density far into the tails, thereby allowing the data to
dominate the posterior more strongly if it disagrees with the prior. We fit our
models with 5,000 total iterations per chain, 2500 warm-up iterations, four
chains, and discarding every second sample. We checked for chain convergence
visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
\(n_\mathrm{eff} >199\) for all parameters). Furthermore, we checked our model
against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
without uncertainty around the predictors.

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
\citep[e.g.~the $t$ distribution compared to a mixture
of two normal distributions;][]{ward2007}.
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution
Furthermore, if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 4$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 4$ allows for an event
three standard deviations below the mean to happen
once every 50 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

>>>>

% TODO 
% change all ln in figures 
\section{Black-swan events in animal populations}

@@

We chose the $t$ distribution for a number of reasons. 
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes 
that could give rise to extreme events in population dynamics,
such as extreme climate events
\citep{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses 
would be important for diagnosing the underlying causes 
of extreme events and to develop predictive models, 
it would be impossible to tease out each of these 
underlying causes from population trend data alone. 
Instead we seek a simple flexible phenomenological model 
that can be applied across the widest possible 
range of population time series to provide a 
comparative overview of the prevalence 
of black-swan events.

One alternative approach would involve fitting 
the generalized extreme value (GEV) distribution,
which represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold.
Therefore, the GEV requires longer time series then typically available
for wild animal population abundance.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
The Gompertz model represents population growth as a linear function in 
log space (here and throughout we use log 
to refer to the natural logarithm log).
If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We also considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet[][p.~443]{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.

@@

\subsubsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals: 
\begin{align} 
x_t &= \lambda + b x_{t-1} + \epsilon_t\\ 
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \citep{connors2014} and the chosen prior in previous
analyses \citep{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was systematically affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

@@

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\log\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{log} abundance.

@@

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from \citet{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \citep[e.g.][]{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

@@

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \citep{gelman2006c,
gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
probability density around expected parameter values while allowing for a
higher probability density far into the tails, thereby allowing the data to
dominate the posterior more strongly if it disagrees with the prior. We fit our
models with 5,000 total iterations per chain, 2500 warm-up iterations, four
chains, and discarding every second sample. We checked for chain convergence
visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
against a similar model with the glmmADMB \textsf{R} package \citep{glmmadmb}
without uncertainty around the predictors.

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
\citep[e.g.~the $t$ distribution compared to a mixture
of two normal distributions;][]{ward2007}.
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution.
Furthermore, if the degrees of freedom parameter ($\nu$)
is fixed,
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 4$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 4$ allows for an event
three standard deviations below the mean to happen
once every 50 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

### 2016-06-25 text/supp.tex

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.2cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{0.4cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\

@@

\caption{Probability density of the Bayesian priors for the Gompertz models. (\textbf{A}) per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. (\textbf{B}) scale parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)$ (\textbf{C}) Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. (\textbf{D}) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{C} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

\caption{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$. Alternative models are (\textbf{A})
    Ricker-logistic; (\textbf{B}) a Gompertz model with AR1 residual correlation;
    (\textbf{C}) modelling population growth rate with no density dependence; and
    (\textbf{D}) a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{B} and \textbf{D}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{B} and four populations in \textbf{D} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. \textbf{A} shows a weaker prior on $\nu$, \textbf{B} shows a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

@@

\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (\textbf{A} to \textbf{C}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), (\textbf{D}) number of time steps, (\textbf{E}) body length, and (\textbf{F}) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{A} and \textbf{D} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}

@@

\caption{Skewness and heavy-tailed parameter posteriors. Top panel shows the shape of the skew-$t$ distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$). Lower panel shows the posterior medians and 50\% credible intervals for the two parameters. The horizontal red line indicates the median ln(skewness) parameter across all populations.}

>>>>

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{6.0cm}>{\RaggedRight}p{0.7cm}>{\RaggedRight}p{1.2cm}>{\RaggedRight}p{4.4cm}>{\RaggedRight}p{1.4cm}}
\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\

@@

\caption{Probability density of the Bayesian priors for the Gompertz models. \textbf{(a)} per capita growth rate at $\log$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$. \textbf{(b)} scale parameter of t-distribution process noise: $\sigma \sim \mathrm{half\mhyphen Cauchy} (0, 2.5)$ \textbf{(c)} Student-$t$ distribution degrees of freedom parameter: $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$. \textbf{(d)} AR1 correlation coefficient of residuals: $\phi \sim \mathrm{truncated \mhyphen Normal}(0, 0.5, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$, the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$. Panel \textbf{(c)} also shows two alternative priors: a weaker prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.005, \mathrm{min.} = 2)$, and a stronger prior $\nu \sim \mathrm{truncated\mhyphen Exponential}(0.02, \mathrm{min.} = 2)$. The inset panel shows the same data but with a log-transformed x axis. Note that the base and weaker priors are relatively flat within the region of $\nu < 20$ with which we are concerned.}

@@

@@

\caption{Estimates of $\nu$ from alternative models plotted against the
    base Gompertz model estimates of $\nu$. Alternative models are \textbf{(a)}
    Ricker-logistic; \textbf{(b)} a Gompertz model with AR1 residual correlation;
    \textbf{(c)} modelling population growth rate with no density dependence; and
    \textbf{(d)} a Gompertz model with fixed magnitude of observation error. Shown
are medians of the posterior (dots) and 50\% credible intervals (segments). The
diagonal line indicates a one-to-one relationship. Different colours indicate
various taxonomic classes. The grey-shaded regions indicate regions of
disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics.
The Gompertz observation error model assumes a fixed standard deviation of
observation error of $0.2$ on a log scale. Models in \textbf{(b)} and \textbf{(d)}
that had not met convergence thresholds were run for four chains, 256,000 iterations (128,000 warm up), and retaining every 10th sample, after which the MCMC chains for one population in \textbf{(b)} and four populations in \textbf{(d)} had still not met convergence thresholds. These populations are excluded from their respective panels, but all populations with Pr($\nu < 10$) $> 0.5$ in the base model had converged and are shown.}

\caption{Estimates of $\nu$ from Gompertz models with alternative priors on $\nu$. \textbf{(a)} shows a weaker prior on $\nu$, \textbf{(b)} shows a stronger prior on $\nu$, as illustrated in Fig.~1. Shown are medians of the posterior (dots) and 50\% credible intervals (segments). The diagonal line indicates a one-to-one relationship. Different colours indicate various taxonomic classes. The grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken as a threshold of heavy-tailed dynamics. In general, the estimates are nearly identical in cases where the data are informative about low values of $\nu$. When the data are less informative about low values of $\nu$, the prior can slightly pull the estimates of $\nu$ towards higher or lower values.}

@@

\caption{Potential covariates of heavy-tailed population dynamics (indicated by a high probability that $\nu < 10$). Shown are (\textbf{a} to \textbf{c}) parameters from the Gompertz heavy-tailed population model ($\sigma$, $\lambda$, $b$), \textbf{(d)} number of time steps, \textbf{(e)} body length, and \textbf{(f)} lifespan. For the Gompertz parameters, $\sigma$ refers to the scale parameter of the Student-$t$ process-noise distribution, $\lambda$ refers to the expected log abundance at the next time step at an abundance of one, $b$ refers to the density dependence parameter. Circles representing a few sharks, crustaceans, and gastropods are filled in white. Median and 90\% credible interval posterior predictions of a beta regression hierarchical model are shown in panels \textbf{(a)} and \textbf{(d)} where there was a high probability the slope coefficient was different from zero (Fig.~3B).}

@@

\caption{Skewness and heavy-tailed parameter posteriors. Top panel shows the shape of the skew-$t$ distribution at different combinations of skewness ($\gamma$) and heavy-tailedness ($\nu$). Lower panel shows the posterior medians and 50\% credible intervals for the two parameters. The horizontal red line indicates the median log(skewness) parameter across all populations.}
